---
tag:
  - DataVis
title: "2020 US Election Results vs Vaccination"
author: "Kazmer Nagy-Betegh"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(fivethirtyeight)
library(here)
library(skimr)
library(janitor)
library(vroom)
library(tidyquant)
library(rvest) # to scrape wikipedia page
```

# Introduction

The purpose of this exercise is to reproduce a plot using your `dplyr` and `ggplot2` skills. Read the  article [The Racial Factor: There's 77 Counties Which Are Deep Blue But Also Low-Vaxx. Guess What They Have In Common?](https://acasignups.net/21/07/18/racial-factor-theres-77-counties-which-are-deep-blue-also-low-vaxx-guess-what-they-have) and have a look at the attached figure.


<img src = "/vaxxes_by_state_red_blue_every_county_070321_1.jpg" />

Datasets that are going to be used for the exercise.

1. To get vaccination by county, we will use [data from the CDC](https://data.cdc.gov/Vaccinations/COVID-19-Vaccinations-in-the-United-States-County/8xkx-amqh) 
1. You need to get [County Presidential Election Returns 2000-2020](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/VOQCHQ)
1. Finally, you also need an estimate of the [population of each county](https://www.ers.usda.gov/webdocs/DataFiles/48747/PopulationEstimates.csv?v=2232)
 


```{r, echo=TRUE, cache=TRUE}

# Download CDC vaccination by county
cdc_url <- "https://data.cdc.gov/api/views/8xkx-amqh/rows.csv?accessType=DOWNLOAD"
vaccinations <- vroom(cdc_url) %>% 
  janitor::clean_names() %>% 
  filter(fips != "UNK") # remove counties that have an unknown (UNK) FIPS code

# Download County Presidential Election Returns
# https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/VOQCHQ -- Download the file from the given URL into your base repository
election2020_results <- vroom(here::here("csv", "countypres_2000-2020.csv")) %>% #read the file from the location you saved it to
  janitor::clean_names() %>% 
  
  # just keep the results for the 2020 election
  filter(year == "2020") %>% 
  
  # change original name county_fips to fips, to be consistent with the other two files
  rename (fips = county_fips)

# Download county population data
population_url <- "https://www.ers.usda.gov/webdocs/DataFiles/48747/PopulationEstimates.csv?v=2232"
population <- vroom(population_url) %>% 
  janitor::clean_names() %>% 
  
  # select the latest data, namely 2019
  select(fips = fip_stxt, pop_estimate_2019) %>% 
  
  # pad FIPS codes with leading zeros, so they are always made up of 5 characters
  mutate(fips = stringi::stri_pad_left(fips, width=5, pad = "0"))

```

### Explore the Data

A quick look at the columns within each dataframe tells us which variables we need to filter for later on to build the graph.

```{r}
head(election2020_results)
head(population)
head(vaccinations)
```


### Data Filtering 

Every datapoint that was below 90% completeness was dropped to be in line with the authors process. The graph would be even closer if the seperate datasets used by the author would be downloaded as well. [Authors Description of Data Utilisation](https://acasignups.net/21/08/31/weekly-update-us-covid19-vaccination-levels-county-partisan-lean1)


```{r}
vax_complete_pop_pct <- vaccinations%>%
  filter(completeness_pct>90.0)%>%
  group_by(fips)%>%
  summarise(series_complete_pop_pct = max(series_complete_pop_pct))
  


population <- population%>%
  mutate(pop = pop_estimate_2019) #add new shorted column name for simpler access

vax_complete_pop_pct <- left_join(x = vax_complete_pop_pct, 
                                  y = population, by = "fips")%>%
  na.omit() #left join the population data table with the vaccination data based on the county codes.


trump_votes <- election2020_results%>%
  filter(candidate == "DONALD J TRUMP")%>% #filter for Donald Trump
  filter(mode == "TOTAL")%>% #only interested in Total votes
  group_by(fips)%>%
  mutate(prcOfVote = (candidatevotes/totalvotes)*100)

head(trump_votes)

vax_complete_pop_pct <- left_join(x = vax_complete_pop_pct, y = trump_votes, by="fips")%>%
                        na.omit()


```

```{r}
library(Hmisc)

ggplot(vax_complete_pop_pct, aes(x = prcOfVote, y = series_complete_pop_pct))+
  geom_point(alpha = 0.2, #set the transparency level
             color = "snow4", # https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf easily find color names
             aes(size = pop/10^6))+ # set the size of the points to scale with the county population size, get population numbers in the millions
  scale_size(range = c(.1, 15), #set the size range for the county circles
             name = "Population (M)",)+ #set the caption
  geom_point(size= 0.1)+ # add the points to the graph again with given size to mark the middle of the county circles
  annotate("rect", #use anotate to only apply changes to one layer of the graph and not all the data in the graph
           xmin = 45, #define the limits of a rectangle
           xmax = Inf,
           ymin = -Inf, 
           ymax = Inf, 
           fill = "indianred1", #specify the color
           alpha = 0.3)+ 
  annotate("rect", 
           xmin = 0, 
           xmax = 55, 
           ymin = -Inf, 
           ymax = Inf, 
           fill = "royalblue1", 
           alpha = 0.3)+
  annotate("line", #add a line for herd immunity
           x = seq(0,100), 
           y = 85, 
           lty = 2, 
           color = "blue")+
  annotate("text", #add herd immunity text
           x = 17, 
           y = 87, 
           label = "Herd Immunity threshold (?)", 
           size = 2, 
           fontface = 4, 
           color = "blue")+
  annotate("text", #see the calculation for the fomula in the last code snippet
           x = 15, 
           y = 15, 
           label = "y = -0.4956x + 0.73669\nR\u00B2 = 0.501", 
           size = 2, 
           color= "red", 
           hjust = "centre", 
           fontface = "bold")+
  annotate("line", 
           x = seq(0,100), 
           y = 51.65, 
           lty = 2,
           color = "blue")+
  annotate("text", 
           x = 10, 
           y = 54,
           label =  sprintf("ACTUAL: %0.2f %%",51.65), 
           size =2, 
           fontface = "bold", 
           color = "blue")+
  annotate("line", 
           x = seq(0,100), 
           y = 70, 
           lty = 2,
           color = "blue")+
  annotate("text", x = 10, y = 72, label =  sprintf("TARGET: %0.2f %%",70), size =2, fontface = "bold", color = "blue")+
  annotate("text", x = 40, y = 15, label = "5/09/2021", color = "red", fontface = "bold", size = 2)+
  geom_smooth(method = "lm", 
              se = FALSE, 
              lty=5, 
              color = "blue", 
              lwd = 0.5)+
  ylab("% of Total Population Vaccinated")+
  xlab("2020 Trump vote %")+
  labs(title = "COVID-19 VACCINATION LEVELS \nOUT OF TOTAL POPULATION BY COUNTY")+
  theme_light()+
  scale_y_continuous(expand = c(0,0), 
                     labels = function(y) paste0(y, "%"), 
                     breaks = scales::pretty_breaks(n=20), 
                     limits = c(0,100))+
  scale_x_continuous(expand = c(0,0), 
                     labels = function(x) paste0(x, "%"), 
                     breaks = scales::pretty_breaks(n=20), 
                     limits = c(0,100))+
  theme(aspect.ratio = 20/18,
        axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title = element_text(size = 10),
        plot.title = element_text(size = 10, 
                                  face = "bold", 
                                  hjust = 0.5),
        legend.title = element_text())
  
  

```

```{r}
print(summary(lm(vax_complete_pop_pct,formula = series_complete_pop_pct ~ prcOfVote)))

sprintf("Percentage of the total population vaccinated: %0.2f ",sum(vax_complete_pop_pct$series_complete_pop_pct*vax_complete_pop_pct$pop)/sum(vax_complete_pop_pct$pop))
```
